A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-06-29 manuscript.qmd
add note on equal author contribution; add johnson citation; add many labs citation and move comment to footnote; update simglm citation; update bibfile

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
date: 2024-06-12
date-format: iso
author:
  - name: Florian Pargent
    equal-contributor: true
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Timo K. Koch
    corresponding: true
    equal-contributor: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - ref: lmu
  - name: Anne-Kathrin Kleine
    orcid: 0000-0003-1919-2834
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    orcid: 0000-0002-6600-9580
    roles:
      - conceptualization
      - funding acquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    orcid: 0000-0002-1633-4772
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: UCL Global Business School for Health
author-note:
  disclosures:
    data-sharing: This is version 2.5 of our preprint published at <https://doi.org/10.31234/osf.io/rpjem>. All materials (reproducible manuscript, including R code and all simulation results) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/dhwf4/>. A Quarto Manuscripts website is hosted at <https://Timo-Ko.github.io/glmm_simulation_tutorial/>.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525).
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. The recent replication crisis underlines the need for more rigorous statistical methodology and adequately powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. However, available methods and software for a priori sample size planning for GLMMs are often limited to specific designs. Tailored data simulation approaches offer a more flexible alternative. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori sample size planning with GLMMs. We not only focus on power analysis but also explain how to use the precision of parameter estimates to determine appropriate sample sizes. We conclude with an outlook on the increasing importance of simulation-based sample size planning.
keywords: [tutorial, sample size planning, generalized linear mixed model, power analysis, precision, data simulation]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

Most empirical studies in psychology and other social sciences apply hypothesis testing. Consequently, the dominant approach for determining an adequate sample size is based on power analysis (i.e., planning for power) [@lakensSampleSizeJustification2022; @maxwellSampleSizePlanning2008]. 
Statistical power is defined as the probability that a HT has a significant p-value when analyzing repeated samples from a population with a true effect of some pre-specified size [@cohenPowerPrimer1992]. Less formally, power is described as the probability that a HT correctly rejects the null hypothesis when the alternative hypothesis is true. If the sample size (i.e., the number of participants and/or stimuli) is insufficient to detect the effects or relationships being investigated with high probability, the study is considered "underpowered".
When planning for power, a target is set for the statistical power of a HT of interest. Assuming an effect size of interest and a desired significance level, a minimum sample size can be determined that, on average, would guarantee reaching this target. 
The most prominent heuristic is to target a power of $1- \beta = 0.8$ in combination with a type I error rate of $\alpha = 0.05$ [@lakensSampleSizeJustification2022]. However, depending on the research goals or resource constraints, there are often good reasons to move away from this standard [@benjaminRedefineStatisticalSignificance2017; @lakensJustifyYourAlpha2018]. For example, the *Social Sciences Replication Project* targeted a power of 0.9 to safeguard against biased effect sizes in the original studies [@camererEvaluatingReplicabilitySocial2018].

@@

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. In Table 1, we review available software packages that can be used to perform power analysis for multilevel models in `r cite_r()`. Power analysis methods can be categorized into formula-based, summary-statistics-based and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on exact formulas to calculate power directly. Summary-statistics-based methods use statistical theory to approximate power based on formula-based methods developed for simple t-tests. Simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically, that is what percentage of simulated datasets produces a significant p-value.
Available formula-based and summary-statistics-based software packages for multilevel models often do not include GLMMs or are limited to simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2023; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021; @kainPracticalGuidePower2015; @greenSIMRPackagePower2016a]. However, many of these tutorials focus on LMMs and the most common study designs [see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers using more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions about the model structure that should align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in @maxwellSampleSizePlanning2008 and @kumleEstimatingPowerGeneralized2021). We will discuss a number of strategies on how to specify model parameters for application-specific, tailored data simulations in a later chapter.

@@

packagetable <- data.frame(
  col1 = c("glmmrBase", "longpower", "mixedpower", "mlmpower", "pamm", "pass.lme", "PowerAnalysisIL", "powerlmm", "simglm", "simr", "summary_statistics based_power", "WebPower"),
  col2 = c("Formula- and simulation-based power analysis; non-linear fixed effects and flexible covariance functions",
              "Formula-based power analysis; focus on longitudinal data",
              "Simulation-based power analysis; based on 'simr' package",
              "Simulation-based power analysis; missing data mechanisms",
              "Simulation-based power analysis; based on 'lme4' package",
              "Formula-based power analysis; limited documentation",
              "Simulation-based power analysis; focus on longitudinal data",
              "Simulation-based power analysis; missing data mechanisms; focus on longitudinal data",
              "Simulation-based power analysis; missing data mechanisms",
              "Simulation-based power analysis; based on 'lme4' package",
              "Summary-statistics-based power analysis; based on 'pwr' package",
              "Formula-based power analysis; based on 'longpower' package; focus on longitudinal data"),
  col3 = c("no", "[yes](https://atrihub.shinyapps.io/power/)", "no", "no", "no", "no", "[yes](https://github.com/ginettelafit/PowerAnalysisIL?tab=readme-ov-file#shiny-app-and-r-package-to-perform-power-analysis-to-select-the-number-of-participants-in-intensive-longitudinal-studies)", "[yes](https://github.com/rpsychologist/powerlmm?tab=readme-ov-file#launch-interactive-web-application)", "[yes](https://simglm.brandonlebeau.org/reference/run_shiny.html)", "no", "[yes](https://koumurayama.shinyapps.io/summary_statistics_based_power/)", "[yes](https://webpower.psychstat.org/wiki/start)"),
  col4 = c("yes", "no", "yes", "no", "no", "no", "no", "no", "yes", "yes", "yes", "no"),
  col5 = c("none", "limited set of study designs with two levels", "none", "limited to two levels", "none", "none", "limited set of study designs with two levels", "limited set of study designs with two or three levels", "limited to two or three levels", "none", "limited set of study designs with two levels", "limited set of study designs with two levels"),
  col6 = c("@watsonGeneralisedLinearMixed2023", "@iddiPowerSampleSize2022", "@kumleEstimatingPowerGeneralized2021", "@endersSimpleMonteCarlo2023", "@martinMeasuringIndividualDifferences2011", "@marcochakyanyuPassLmePower2019", "@lafitSelectionNumberParticipants2021", "@magnussonConsequencesIgnoringTherapist2018", "@lebeauSimglmSimulateModels2017", "@greenSIMRPackagePower2016", "@murayamaSummarystatisticsbasedPowerAnalysis2022", "@zhangPracticalStatisticalPower2018")
)

>>>>

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
date: 2024-06-12
date-format: iso
author:
  - name: Florian Pargent
    equal-contributor: true
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Timo K. Koch
    corresponding: true
    equal-contributor: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - ref: lmu
  - name: Anne-Kathrin Kleine
    orcid: 0000-0003-1919-2834
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    orcid: 0000-0002-6600-9580
    roles:
      - conceptualization
      - funding acquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    orcid: 0000-0002-1633-4772
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: UCL Global Business School for Health
author-note:
  disclosures:
    data-sharing: This is version 2.5 of our preprint published at <https://doi.org/10.31234/osf.io/rpjem>. All materials (reproducible manuscript, including R code and all simulation results) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/dhwf4/>. A Quarto Manuscripts website is hosted at <https://Timo-Ko.github.io/glmm_simulation_tutorial/>.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525).
    gratitude: ~
    authorship-agreements: Florian Pargent and Timo Koch contributed equally to this manuscript.
language:
  title-block-role-introduction: "Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. The recent replication crisis underlines the need for more rigorous statistical methodology and adequately powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. However, available methods and software for a priori sample size planning for GLMMs are often limited to specific designs. Tailored data simulation approaches offer a more flexible alternative. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori sample size planning with GLMMs. We not only focus on power analysis but also explain how to use the precision of parameter estimates to determine appropriate sample sizes. We conclude with an outlook on the increasing importance of simulation-based sample size planning.
keywords: [tutorial, sample size planning, generalized linear mixed model, power analysis, precision, data simulation]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

Most empirical studies in psychology and other social sciences apply hypothesis testing. Consequently, the dominant approach for determining an adequate sample size is based on power analysis (i.e., planning for power) [@lakensSampleSizeJustification2022; @maxwellSampleSizePlanning2008]. 
Statistical power is defined as the probability that a HT has a significant p-value when analyzing repeated samples from a population with a true effect of some pre-specified size [@cohenPowerPrimer1992]. Less formally, power is described as the probability that a HT correctly rejects the null hypothesis when the alternative hypothesis is true. If the sample size (i.e., the number of participants and/or stimuli) is insufficient to detect the effects or relationships being investigated with high probability, the study is considered "underpowered".
When planning for power, a target is set for the statistical power of a HT of interest. Assuming an effect size of interest and a desired significance level, a minimum sample size can be determined that, on average, would guarantee reaching this target. 
The most prominent heuristic is to target a power of $1- \beta = 0.8$ in combination with a type I error rate of $\alpha = 0.05$ [@lakensSampleSizeJustification2022]. However, depending on the research goals or resource constraints, there are often good reasons to move away from this standard [@benjaminRedefineStatisticalSignificance2017; @lakensJustifyYourAlpha2018].^[For example, the *Social Sciences Replication Project* targeted a power of 0.90 to safeguard against biased effect sizes in the original studies [@camererEvaluatingReplicabilitySocial2018], and *Many Labs 5* even targeted a power of 0.95 [@ebersoleManyLabsTesting2020].]

@@

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. In Table 1, we review available software packages that can be used to perform power analysis for multilevel models in `r cite_r()`. Power analysis methods can be categorized into formula-based, summary-statistics-based and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on exact formulas to calculate power directly. Summary-statistics-based methods use statistical theory to approximate power based on formula-based methods developed for simple t-tests. Simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically, that is what percentage of simulated datasets produces a significant p-value.
Available formula-based and summary-statistics-based software packages for multilevel models often do not include GLMMs or are limited to simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2023; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021; @kainPracticalGuidePower2015; @greenSIMRPackagePower2016a; @johnsonPowerAnalysisGeneralized2015]. However, many of these tutorials focus on LMMs and the most common study designs [see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers using more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions about the model structure that should align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in @maxwellSampleSizePlanning2008 and @kumleEstimatingPowerGeneralized2021). We will discuss a number of strategies on how to specify model parameters for application-specific, tailored data simulations in a later chapter.

@@

packagetable <- data.frame(
  col1 = c("glmmrBase", "longpower", "mixedpower", "mlmpower", "pamm", "pass.lme", "PowerAnalysisIL", "powerlmm", "simglm", "simr", "summary_statistics based_power", "WebPower"),
  col2 = c("Formula- and simulation-based power analysis; non-linear fixed effects and flexible covariance functions",
              "Formula-based power analysis; focus on longitudinal data",
              "Simulation-based power analysis; based on 'simr' package",
              "Simulation-based power analysis; missing data mechanisms",
              "Simulation-based power analysis; based on 'lme4' package",
              "Formula-based power analysis; limited documentation",
              "Simulation-based power analysis; focus on longitudinal data",
              "Simulation-based power analysis; missing data mechanisms; focus on longitudinal data",
              "Simulation-based power analysis; missing data mechanisms",
              "Simulation-based power analysis; based on 'lme4' package",
              "Summary-statistics-based power analysis; based on 'pwr' package",
              "Formula-based power analysis; based on 'longpower' package; focus on longitudinal data"),
  col3 = c("no", "[yes](https://atrihub.shinyapps.io/power/)", "no", "no", "no", "no", "[yes](https://github.com/ginettelafit/PowerAnalysisIL?tab=readme-ov-file#shiny-app-and-r-package-to-perform-power-analysis-to-select-the-number-of-participants-in-intensive-longitudinal-studies)", "[yes](https://github.com/rpsychologist/powerlmm?tab=readme-ov-file#launch-interactive-web-application)", "[yes](https://simglm.brandonlebeau.org/reference/run_shiny.html)", "no", "[yes](https://koumurayama.shinyapps.io/summary_statistics_based_power/)", "[yes](https://webpower.psychstat.org/wiki/start)"),
  col4 = c("yes", "no", "yes", "no", "no", "no", "no", "no", "yes", "yes", "yes", "no"),
  col5 = c("none", "limited set of study designs with two levels", "none", "limited to two levels", "none", "none", "limited set of study designs with two levels", "limited set of study designs with two or three levels", "limited to two or three levels", "none", "limited set of study designs with two levels", "limited set of study designs with two levels"),
  col6 = c("@watsonGeneralisedLinearMixed2023", "@iddiPowerSampleSize2022", "@kumleEstimatingPowerGeneralized2021", "@endersSimpleMonteCarlo2023", "@martinMeasuringIndividualDifferences2011", "@marcochakyanyuPassLmePower2019", "@lafitSelectionNumberParticipants2021", "@magnussonConsequencesIgnoringTherapist2018", "@lebeauPowerAnalysisSimulation2019", "@greenSIMRPackagePower2016", "@murayamaSummarystatisticsbasedPowerAnalysis2022", "@zhangPracticalStatisticalPower2018")
)
names(packagetable) <- c("Package name", "Summary", "GUI available", "GLMM support", "Design limitations", "Reference")
